Homicide Increases in the COVID-19 Era

### 2025-03-08 Draft3.qmd
Submitted version

author: "Marco Stine, Kevin Hu, Bianca Schutz, and Kaitlyn Kirt"
fontsize: 10pt
toc: true                   # add table of contents at the beginning
toc-depth: 3               
geometry: margin=1in
bibliography: references.bib
format:
  pdf: default
number-sections: true
---

@@

[ADD MAIN EDA TAKEAWAYS HERE]

## Examining Crime Density from 2010–2019 and 2020-Present

```{r, echo=FALSE, warning=FALSE, message=FALSE}
lapd_shape <- read_sf(dsn = "0_Data/Raw_Data/LAPD_Divisions", layer = "LAPD_Divisions")

@@

smoothScatter(
  x    = crime_la_2010_2019$LON,
  y    = crime_la_2010_2019$LAT,
  xlim = c(-119, -118),
  ylim = c(33.5, 34.5),
  nbin = 300,
  xlab = "Longitude",
  ylab = "Latitude",
  main = "Crime Density by LAPD Division (2010–2019)",
  colramp = colorRampPalette(c("white", "skyblue", "blue", "darkblue")),
  cex.main=0.8
)

@@

smoothScatter(
  x    = crime_la_2020_present$LON,
  y    = crime_la_2020_present$LAT,
  xlim = c(-119, -118),
  ylim = c(33.5, 34.5),
  nbin = 300,
  xlab = "Longitude",
  ylab = "Latitude",
  main = "Crime Density by LAPD Division (2020–Present)",
  colramp = colorRampPalette(c("white", "pink", "red", "darkred")),
  cex.main=0.8
)

@@

### Interpretation

Both time periods show major clusters of reported crimes in central LA. The 2020–present data appears slightly more dispersed to the north, but overall patterns remain similar looking.

## Examining Number Of Crimes Over Time

```{r, echo=FALSE, warning=FALSE, message=FALSE}
crime_la$DateParsed <- as.POSIXct(crime_la$Date.Rptd, 
                                  format = "%m/%d/%Y %I:%M:%S %p", 
                                  tz = "America/Los_Angeles")

@@

plot(
  weekly_counts$Index,
  weekly_counts$Crimes,
  type = "l",
  col  = "blue",
  lwd  = 2,
  xlab = "Weekly Index",
  ylab = "Number of Crimes",
  main = "Weekly Crime Counts in Los Angeles (2010–Present)"
)

@@

### Interpretation

The time series shows stable weekly crime counts with occasional large spikes. These surges indicate certain weeks where crime reporting jumped significantly, there is a recent interesting surge that seems to mean-revert in 2024.

## Crime rate frequency by day of the week

```{r, echo=FALSE, fig.cap = "Crime Occurrences by Day of the Week", warning=FALSE, message=FALSE}
# Convert Date.Rptd to Date format (ignore time part)
#crime_la$Date.Rptd
crime_la$Date <- as.Date(substr(crime_la$Date.Rptd, 1, 10), format = "%m/%d/%Y")

@@

# Order the days correctly
day_order <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
crime_counts <- crime_counts[day_order]

# Plot bar chart with horizontal labels
barplot(crime_counts, col = "lightblue",
        xlab = "Day of the Week", ylab = "Number of Crimes",
        names.arg = names(crime_counts), las = 1, cex.names = 0.4, cex.axis = 0.6)
```

### Interpretation

The number of crimes are rather evenly distributed, with the weekends having slightly lower crime compared to week days.

## Gender Distribution for crimes over time

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Remove missing values and empty strings in Vict.Sex column
crime_la <- subset(crime_la, Vict.Sex != "" & !is.na(Vict.Sex))

@@

# Create a pie chart using base R
color <- c("F" = "pink", "M" = "lightblue", "N" = "purple")
pie(
  gender_counts, 
  labels = paste(names(gender_counts), 
                 round(gender_counts / sum(gender_counts) * 100, 1), "%"), 
  col = color[names(gender_counts)],
  main = "Distribution of Crime Victims by Gender"
)
```

### Interpretation

From the pie chart, we can see that around 50% of the crime is commited by male, 45% by female and 5% by non-binary.

## Median Household Income by Year and Zip Code in Los Angeles

```{r, echo=FALSE, warning=FALSE, message=FALSE}
zcta <- read_sf(dsn = "0_Data/Raw_Data/cb_2020_us_zcta520_500k", layer = "cb_2020_us_zcta520_500k")
LA_zctas <- read.csv("0_Data/Raw_Data/zctas.csv") #obtained using zipcodeR::search_city("Los Angeles", "CA") and then checking which zip codes are also ZCTAs

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# quantiles for the entire dataset (from 2018 to 2023) to divide the median income into groups
qntls <- round(quantile(zcta_no_NAs$S1903_C03_001E, probs = seq(0, 1, length.out = 7)), 0)

@@

### Interpretation

The maps demonstrate that median income in certain LA zip codes (particularly those to the west) have a much higher median household income than the rest of the city, but over the last six years, the median income in most other LA zip codes has increased. However, there are some that a typical family is at or below the poverty line. We intend to investigate if there is a spatial correlation between these poorer zip codes and the incidence of crimes, particularly homicide, and whether this is affected by pandemic conditions.

## Crime Rates Compared to Median Household Income

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# using latitude and longitude to obtain the zip codes of the crime locations
# crime_la$year <- as.integer(stringr::str_sub(crime_la$YearWeek, 1, 4))
# id_lat_long_crime_la <- subset(crime_la, select = c("LAT", "LON", "DR_NO", "year"))

# taking a random sample to reduce load on model
#set.seed(123)
#sampled_ids <- id_lat_long_crime_la[sample(1:nrow(id_lat_long_crime_la), 500),]
#rownames(sampled_ids) <- NULL
#zip_codes <- reverse_geocode(sampled_ids, lat = LAT, long = LON, verbose = TRUE)
#write.csv(zip_codes, "sampled_crimes.csv")
zip_codes <- read.csv("0_Data/Cleaned_Data/sampled_crimes.csv")

zip_codes$ZCTA5CE20 <- stringr::str_extract(zip_codes$address, "\\d{5}")

crime_income_sample <- zip_codes %>% inner_join(unique(subset(zcta_no_NAs, select = c("ZCTA5CE20", "S1903_C03_001E", "year"))), by = c("ZCTA5CE20", "year"))

# this takes a while, so loading in this csv instead

crime_income_sample <- zip_codes %>% inner_join(unique(subset(zcta_no_NAs, select = c("ZCTA5CE20", "S1903_C03_001E", "year"))), by = c("ZCTA5CE20", "year"))

crime_income_per_zip <- crime_income_sample %>% 
  group_by(ZCTA5CE20) %>% 
  summarize(number_of_crimes = n(),
            cali_median_income = mean(S1903_C03_001E)) %>%
  ungroup()

plot(x = crime_income_per_zip$cali_median_income, 
     y = crime_income_per_zip$number_of_crimes, 
     type = "p",
     ylab = "Number of Crimes",  
     xlab = "Median Income ($)",  
     bty = "l")
```

### Interpretation

Based on this plot, we do not observe a clear trend between the number of crimes committed and the median income in that zip code. However, this scatterplot only accounts for a small portion of the dataset and averages across pre- and post-pandemic conditions. Future visualizations and analyses could consider pandemic vs non-pandemic years as variables.

# Exploratory Data Analysis With ggplot2

## Monthly Frequency of Evictions and Crimes in the City of Los Angeles, February 2023 to February 2025

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# fix mistake
evictions_la$Date.Received <- gsub("2205", "2025", evictions_la$Date.Received)

@@

### Interpretation

This histogram plots the frequency of evictions in Los Angeles from February 2023 to February 2025, where the City of Los Angeles Housing Department has published case data. By summarizing both the crime and the evictions datasets by month, we see that the number of evictions was highest in early-mid 2023, and since, there have typically been slightly lower frequencies of evictions. Additionally, we split the number of crimes per month based on the average number of crimes of 14,439 to denote above and below average number of crimes per month. There is a clear split between March and April of 2024, as all months before that had above average crime, and all after had below average. Interestingly, the months where there is below average crime, seem to be the ones with slightly lower numbers of evictions.

## Annual Number of Crimes by Los Angeles Police Department Bureau

```{r, echo=FALSE, warning=FALSE, message=FALSE}
crime_la$year <- lubridate::year(crime_la$DateParsed)
crime_by_area <- crime_la %>% group_by(AREA.NAME, year) %>% summarize(numberOfCrimes = n()) %>% ungroup()

@@

### Interpretation

This boxplot shows the number of crimes in the four different bureaus of Los Angeles annually. We categorized the areas into their respective Central, South, Valley, and West Bureaus. The South Bureau has the highest median number of crimes, as well as the largest variability, while the Central, Valley, and West bureaus have relatively similar distributions with lower crime counts.The outliers suggests some years had significantly lower crime counts than others. This visualization helps us understand how crime is distributed across different regions, with the South experiencing the highest fluctuation and overall crime levels.

## Correlations Between the Number of Crimes and Different Law Enforcement and Socioeconomic Factors

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(lubridate)
# creating dataset for correlations

@@

### Interpretation

This heat map shows the correlations between different variables that could affect the number of crimes committed. We see that the most significant correlations for the number of crimes are the unemployment rate, year, and number of evictions. Years is positively correlated with number of crimes, meaning that as time passed, the number of crimes increased. Similarly, as the number of evictions increases, so does the number of crimes. Interestingly, as the unemployment rate increases, the number of crimes decreases, which is the opposite effect of what was expected (one would anticipate that higher unemployment would result in more crime). The LAPD budget interestingly had a medium-strength positive correlation with number of crimes, suggesting we should further examine the increase in the LAPD budget compared to the increase in crimes.

## Homicide Victims by Age and Gender in Los Angeles 2010-Present

```{r, echo=FALSE, warning=FALSE, message=FALSE}
homicide_by_gender <- crime_la %>% filter(Crm.Cd.Desc == "CRIMINAL HOMICIDE") %>% dplyr::select(Vict.Sex, DateParsed, Vict.Age) %>% mutate(gender = factor(Vict.Sex))

@@

### Interpretation

This jitter plot shows us both the frequencies and the demographics of homicide victims in Los Angeles from 2010 to present. Based on the density of the points, we see that most homicide victims in the last 15 years have been men between the ages of around 15 to 65. Additionally, we see that there are not many victims who are toddlers and younger children, but there is a good number of babies who have been killed. This is an intricacy in our data that might be interesting to look into further, particularly into the circumstances of those deaths and potential correlations between reports of child abuse and/or domestic violence and infanticide.

## Frequency of Crimes in Los Angeles Throughout the Day

```{r, echo=FALSE, warning=FALSE, message=FALSE}
crime_la$time <- sprintf("%04d",crime_la$TIME.OCC)
crime_la$Hour <- as.numeric(substr(as.character(crime_la$time), start = 1, stop = 2))

@@

### Interpretation

This density plot shows the frequency of crimes depending on the hour (in military time) that they occurred. Based on the plot, we see that fewer crimes occur early in the morning (with the lowest at 5 AM). The data peaks at noon, meaning the most crimes occur in that hour of the day. Crimes are more frequent after that, until the early morning. The second most common crime occurrence is 6 PM, or 18 hours.

## Time series analyzing homicide rate trends over time by socioeconomic quartiles

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Data Preparation
homicides <- crime_la %>% inner_join(crimes_with_zips, by = c("DR_NO", "LAT", "LON","X", "year")) %>% filter(Crm.Cd.Desc == "CRIMINAL HOMICIDE") %>% 
  mutate(date = as.Date(DateParsed),
         zip = as.character(ZCTA5CE20)) %>% dplyr::select(-X)

@@

# creating labels
levels(homicide_trends$income_quartile) <- c("Low Income", "Lower-Middle Income", "Higher-Middle Income", "High Income")
# Plot homicide trends over time by income quartile
ggplot(homicide_trends, aes(x = year, y = homicide_count)) +
  geom_area(aes(fill = as.factor(income_quartile)), alpha = 0.6, position = "identity") +
  labs(
    x = "Year",
    y = "Number of Homicides",
    fill = "Income Quartile",
    title = "Homicide Rates in Los Angeles By Median Income Level"
  ) +
  theme_minimal() +
  scale_fill_viridis_d(option = "plasma", direction = -1)
```

### Interpretation

Homicide rates were low pre-pandemic and had a sharp increase since the pandemic, except in areas of low income, where homicide rates were high before the pandemic. In lower income areas, the rates of homicide have decreased significantly since the end of the pandemic. On the other hand, areas of higher-middle and high income have seen large increases in homicide in the six years shown.

[BASED ON THESE PLOTS, WE WILL DO XYZ TO EXAMINE ABC]

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(broom)

@@

ar1_trend_regression %>%
tidy() %>%
  mutate(
    p.value = scales::pvalue(p.value)
  ) %>%
  knitr::kable(
    caption = "Linear Regression Model Results Estimating Crimes Per Week",
    col.names = c("Predictor", "B", "SE", "t", "p"),
    digits = c(0, 2, 3, 2, 3),
    align = c("l", "r", "r", "r", "r")
  )
```

@@

## Regression Between Median Income and Crimes per Zip Code - DROP THIS

```{r, echo=FALSE, warning=FALSE, message=FALSE}
lm_income_crimes <- lm(number_of_crimes ~ cali_median_income, data = crime_income_per_zip)

@@

lm_income_crimes %>%
tidy() %>%
  mutate(
    p.value = scales::pvalue(p.value)
  ) %>%
  knitr::kable(
    caption = "Linear Regression Model Results Estimating Crimes Based on Median Income",
    col.names = c("Predictor", "B", "SE", "t", "p"),
    digits = c(0, 2, 3, 2, 3),
    align = c("l", "r", "r", "r", "r")
  )
```

@@

```{r, warning=FALSE, message=FALSE, echo=FALSE}
library(kableExtra)
crimes_by_cb <- read.csv("0_Data/Cleaned_Data/crimes_by_census_block.csv")
colnames(crimes_by_cb) <- c(colnames(crimes_by_cb)[-c(1, length(colnames(crimes_by_cb)))],"LON", "LAT")

@@

summary(model)
```
### Discussion

## Model Progression
### Initial Spatial Analysis of Crime in Los Angeles

```{r, echo=FALSE, warning=FALSE, message=FALSE}
homicides_latlon_rounded <- homicides %>% mutate(LAT_rounded = round(LAT, 1),
                                                 LON_rounded = round(LON, 1))
homicides_lat_long <- homicides_latlon_rounded %>% group_by(LAT_rounded, LON_rounded) %>% summarize(HomicideCount = n()) %>% ungroup()

linear_model1 <- lm(HomicideCount ~ LAT_rounded + LON_rounded, data = homicides_lat_long)

summary(linear_model1) # longitude rounded is significant, LAT_rounded is not
```

### By Year
```{r, echo=FALSE, warning=FALSE, message=FALSE, eval=FALSE}
homicides_lat_long <- homicides_latlon_rounded %>% group_by(LAT_rounded, LON_rounded) %>% summarize(HomicideCount = n()) %>% ungroup()

linear_model2 <- lm(HomicideCount ~ LAT_rounded + LON_rounded + year, data = homicides_by_year_lat_long)

summary(linear_model2) # longitude rounded is significant, LAT_rounded is not. Neither is year
```

@@

```{r Wrangling Main, eval=FALSE}
library(sf) # for map
library(tidyverse) # using only for joining datasets to build map visual
library(RColorBrewer) # for map color scheme
library(tidygeocoder) # to get zip codes of addresses

# Primary datasets: Crime Data 2020-Present and 2010-2019
crime_la_2010_2019 <- read.csv("0_Data/Raw_Data/Crime_Data_from_2010_to_2019.csv")

@@

# Combine the two datasets
crime_la <- rbind(crime_la_2010_2019, crime_la_2020_present)
```

@@

#### Figure 1 - Examining Crime Density from 2010–2019 and 2020-Present {.unnumbered}

```{r Figure 1 Code, eval=FALSE}

@@

#### Figure 2 - Examining Number Of Crimes Over Time {.unnumbered}

```{r Figure 2 Code, eval=FALSE}
crime_la$DateParsed <- as.POSIXct(crime_la$Date.Rptd, 
                                  format = "%m/%d/%Y %I:%M:%S %p", 
                                  tz = "America/Los_Angeles")

@@

#### Figure 3 - Crime rate frequency by day of the week {.unnumbered}

```{r Figure 3 Code, eval=FALSE}
# Convert Date.Rptd to Date format (ignore time part)
#crime_la$Date.Rptd
crime_la$Date <- as.Date(substr(crime_la$Date.Rptd, 1, 10), format = "%m/%d/%Y")

# Set locale to English (for weekday names)
Sys.setlocale("LC_TIME", "C")

@@

# Order the days correctly
day_order <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
crime_counts <- crime_counts[day_order]

# Plot bar chart with horizontal labels
barplot(crime_counts, col = "lightblue",
        xlab = "Day of the Week", ylab = "Number of Crimes",
        names.arg = names(crime_counts), las = 1, cex.names = 0.8, cex.axis = 0.6)
```

#### Figure 4 - Gender Distribution for crimes over time {.unnumbered}

>>>>

author: "Bianca Schutz, Kaitlyn Kirt, Kevin Hu, and Marco Stine"
fontsize: 10pt
toc: true                   # add table of contents at the beginning
toc-depth: 3               
geometry: margin=1in
bibliography: references.bib
format:
  pdf: default
number-sections: true
---

@@

```{r echo=FALSE, fig.cap="Crime densities in Los Angeles by LAPD Division, message=FALSE, warning=FALSE, 2010.2019 and 2020.Present", fig.align="center"}
lapd_shape <- read_sf(dsn = "0_Data/Raw_Data/LAPD_Divisions", layer = "LAPD_Divisions")

@@

@@

@@

Figure 1 shows that both time periods show major clusters of reported crimes in central LA. The 2020–present data appears slightly more dispersed to the north, but overall, the density of crimes has remained the same over time. The West LA and Foothill LAPD divisions seem to have the lowest density of crimes. West LA Division has crimes mostly located in the south of its area. Beverly Hills, which is not one of the LAPD divisions and has its own police force, explains the whiter spot in central LA.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap = "Weekly Crime Counts in Los Angeles from 2010 to present", fig.align = "center"}
crime_la$DateParsed <- as.POSIXct(crime_la$Date.Rptd, 
                                  format = "%m/%d/%Y %I:%M:%S %p", 
                                  tz = "America/Los_Angeles")

@@

plot(
  weekly_counts$Index,
  weekly_counts$Crimes,
  type = "l",
  col  = "blue",
  lwd  = 2,
  xlab = "Weekly Index",
  ylab = "Number of Crimes")

@@

Figure 2 analyzes the number of crimes over time using time series. The time series shows stable weekly crime counts with occasional large spikes. These surges indicate certain weeks where crime reporting jumped significantly, there is a recent interesting surge that seems to mean-revert in 2024.

```{r, echo=FALSE, fig.cap = "Los Angeles Crimes by Day of the Week", warning=FALSE, message=FALSE, fig.align = "center"}
# Convert Date.Rptd to Date format (ignore time part)
#crime_la$Date.Rptd
crime_la$Date <- as.Date(substr(crime_la$Date.Rptd, 1, 10), format = "%m/%d/%Y")

@@

# Order the days correctly
day_order <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
crime_counts <- data.frame(crime_counts[day_order])
colnames(crime_counts) <- c("Day", "CrimeCount")

crime_counts$thousands <- crime_counts$CrimeCount/1000

# Plot bar chart with horizontal labels
ggplot(crime_counts, aes(x = Day, y = thousands)) +
  geom_col(fill = "blue4") +
  labs(x = "Day of the Week",
       y = "Number of Crimes (thousands)")
```

In Figure 3, the number of crimes are rather evenly distributed, with the weekends having slightly lower crime compared to week days, which may be a result from more people around during weekends which indirectly supervises people's behavior. Most crimes occur on Mondays, and the fewest crimes occur on Sunday. The plot analyzes the crime rate frequency frequency by day of the week using a bar graph.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Percent of crime victims by gender, 2010 to present", fig.align = "center"}
# Remove missing values and empty strings in Vict.Sex column
crime_la <- subset(crime_la, Vict.Sex != "" & !is.na(Vict.Sex))

@@

# Create a pie chart using base R
color <- c("F" = "pink", "M" = "lightblue", "N" = "purple")
pie(
  gender_counts, 
  labels = paste(names(gender_counts), 
                 round(gender_counts / sum(gender_counts) * 100, 1), "%"), 
  col = color[names(gender_counts)]
)
```

From the pie chart shown in Figure 4, we can see that around 50% of crime victims are male, 45% are female and 5% are non-binary. This shows that the distribution of crime victims by gender is relatively even, though men do account for a plurality of the victims.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center"}
zcta <- read_sf(dsn = "0_Data/Raw_Data/cb_2020_us_zcta520_500k", layer = "cb_2020_us_zcta520_500k")
LA_zctas <- read.csv("0_Data/Raw_Data/zctas.csv") #obtained using zipcodeR::search_city("Los Angeles", "CA") and then checking which zip codes are also ZCTAs

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Median household income by year and by ZIP Code Tabulation Area (ZCTA) in Los Angeles", fig.align = "center"}
# quantiles for the entire dataset (from 2018 to 2023) to divide the median income into groups
qntls <- round(quantile(zcta_no_NAs$S1903_C03_001E, probs = seq(0, 1, length.out = 7)), 0)

@@

The maps in Figure 5 demonstrate that median income in certain LA zip codes (particularly those to the west) have a much higher median household income than the rest of the city, but over the last six years, the median income in most other LA zip codes has increased. However, there are some that a typical family is at or below the poverty line. We intend to investigate if there is a spatial correlation between these poorer zip codes and the incidence of crimes, particularly homicide, and whether this is affected by pandemic conditions.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Annual Median Income and Homicide Count by ZCTA in Los Angeles with top 1% income and top 1% homicide rates labelled", fig.align = "center"}
cali_median_income$ZCTA5CE20 <- as.integer(cali_median_income$ZCTA)
crime_income <- crimes_with_zips %>% inner_join(cali_median_income, by = c("year", "ZCTA5CE20"))

crime_type <- crime_la %>% dplyr::select(year, Crm.Cd.Desc, DR_NO)

crime_income_type <- crime_income %>% inner_join(crime_type, by = c("year", "DR_NO"))

homicide_income_by_year <- crime_income_type %>% 
  filter(Crm.Cd.Desc == "CRIMINAL HOMICIDE") %>%
  group_by(year, ZCTA5CE20) %>%
  summarize(HomicideCount = n(),
            cali_median_income = median(as.numeric(S1903_C03_001E))) %>%
  ungroup()

homicide_income_by_year <- homicide_income_by_year %>% filter(!is.na(HomicideCount) & !is.na(cali_median_income))

homicide_income_by_year$label <- ifelse(
  homicide_income_by_year$cali_median_income > quantile(homicide_income_by_year$cali_median_income, 0.99) | 
  homicide_income_by_year$HomicideCount > quantile(homicide_income_by_year$HomicideCount, 0.99), 
  homicide_income_by_year$ZCTA5CE20, 
  "")

ggplot(data = homicide_income_by_year, aes(x = cali_median_income, 
                                        y = HomicideCount)) +
  geom_point(aes(color = as.factor(year))) +
  ggthemes::scale_color_stata() +
  ggrepel::geom_text_repel(aes(label = label), size = 2.5, nudge_x = 2, nudge_y = 1) + 
  labs(x = "Median Annual Income ($)",
       y = "Annual Homicide Count",
       color = "Year")
```

Based on Figure 6, we observe a weak inverse relationship between median annual income and homicide count. Specifically, we see that higher homicide rates occur in ZCTA areas with lower median annual income, though most observations are still near or at zero, meaning few homicides typically occur. However, there are no observations with a median annual income of over \$100,000 where the homicide rate is greater than 10, whereas there are several of those observations for ZCTA/year combinations under \$100,000. The zip code which stands out most on this figure is 90003, which has 3 of the 4 highest homicide counts in the data. This suggests that there might be a relationship between median income and homicide count, though the relationship might be non-linear or weak, suggesting there are other factors at play.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center", fig.cap="Monthly Frequency of Evictions and Crimes in the City of Los Angeles, February 2023 to February 2025"}
# fix mistake
evictions_la$Date.Received <- gsub("2205", "2025", evictions_la$Date.Received)

@@

Figure 7 plots the frequency of evictions in Los Angeles from February 2023 to February 2025, where the City of Los Angeles Housing Department has published case data. By summarizing both the crime and the evictions datasets by month, we see that the number of evictions was highest in early-mid 2023, and since, there have typically been slightly lower frequencies of evictions. Additionally, we split the number of crimes per month based on the average number of crimes of 14,439 to denote above and below average number of crimes per month. There is a clear split between March and April of 2024, as all months before that had above average crime, and all after had below average. Interestingly, the months where there is below average crime, seem to be the ones with slightly lower numbers of evictions.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center", fig.cap="Annual Number of Crimes by Los Angeles Police Department Bureau"}
crime_la$year <- lubridate::year(crime_la$DateParsed)
crime_by_area <- crime_la %>% group_by(AREA.NAME, year) %>% summarize(numberOfCrimes = n()) %>% ungroup()

@@

Figure 8 shows the number of crimes in the four different bureaus of Los Angeles annually. We categorized the areas into their respective Central, South, Valley, and West Bureaus. The South Bureau has the highest median number of crimes, as well as the largest variability, while the Central, Valley, and West bureaus have relatively similar distributions with lower crime counts.The outliers suggests some years had significantly lower crime counts than others. This visualization helps us understand how crime is distributed across different regions, with the South experiencing the highest fluctuation and overall crime levels.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center", fig.cap="Strength of Correlations Between the Number of Crimes and Different Law Enforcement and Socioeconomic Factors"}
library(lubridate)
# creating dataset for correlations

@@

Figure 9 shows the correlations between different variables that could affect the number of crimes committed. We see that the most significant correlations for the number of crimes are the unemployment rate, year, and number of evictions. Years is positively correlated with number of crimes, meaning that as time passed, the number of crimes increased. Similarly, as the number of evictions increases, so does the number of crimes. Interestingly, as the unemployment rate increases, the number of crimes decreases, which is the opposite effect of what was expected (one would anticipate that higher unemployment would result in more crime). The LAPD budget interestingly had a medium-strength positive correlation with number of crimes, suggesting we should further examine the increase in the LAPD budget compared to the increase in crimes.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center", fig.cap="Homicide Victims by Age and Gender in Los Angeles 2010-Present"}
homicide_by_gender <- crime_la %>% filter(Crm.Cd.Desc == "CRIMINAL HOMICIDE") %>% dplyr::select(Vict.Sex, DateParsed, Vict.Age) %>% mutate(gender = factor(Vict.Sex))

@@

The jitter plot in Figure 10 shows us both the frequencies and the demographics of homicide victims in Los Angeles from 2010 to present. Based on the density of the points, we see that most homicide victims in the last 15 years have been men between the ages of around 15 to 65. Additionally, we see that there are not many victims who are toddlers and younger children, but there is a good number of babies who have been killed. This is an intricacy in our data that might be interesting to look into further, particularly into the circumstances of those deaths and potential correlations between reports of child abuse and/or domestic violence and infanticide.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center", fig.cap="Frequency of Crimes in Los Angeles Throughout the Day"}
crime_la$time <- sprintf("%04d",crime_la$TIME.OCC)
crime_la$Hour <- as.numeric(substr(as.character(crime_la$time), start = 1, stop = 2))

@@

The density plot in Figure 11 shows the frequency of crimes depending on the hour (in military time) that they occurred. Based on the plot, we see that fewer crimes occur early in the morning (with the lowest at 5 AM). The data peaks at noon, meaning the most crimes occur in that hour of the day. Crimes are more frequent after that, until the early morning. The second most common crime occurrence is 6 PM, or 18 hours.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center", fig.cap="Time series analyzing homicide rate trends over time by median income quartiles"}
# Data Preparation
homicides <- crime_la %>% inner_join(crimes_with_zips, by = c("DR_NO", "LAT", "LON","X", "year")) %>% filter(Crm.Cd.Desc == "CRIMINAL HOMICIDE") %>% 
  mutate(date = as.Date(DateParsed),
         zip = as.character(ZCTA5CE20)) %>% dplyr::select(-X)

@@

# creating labels
levels(homicide_trends$income_quartile) <- c("Low Income", "Lower-Middle Income", "Higher-Middle Income", "High Income")
# Plot homicide trends over time by income quartile
ggplot(homicide_trends, aes(x = year, y = homicide_count)) +
  geom_area(aes(fill = as.factor(income_quartile)), alpha = 0.6, position = "identity") +
  labs(
    x = "Year",
    y = "Number of Homicides",
    fill = "Income Quartile"
  ) +
  theme_minimal() +
  scale_fill_viridis_d(option = "plasma", direction = -1)
```

Figure 12 shows that homicide rates were low pre-pandemic and had a sharp increase since the pandemic, except in areas of low income, where homicide rates were high before the pandemic. In lower income areas, the rates of homicide have decreased significantly since the end of the pandemic. On the other hand, areas of higher-middle and high income have seen large increases in homicide in the six years shown.

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center"}
library(broom)

@@

sjPlot::tab_model(ar1_trend_regression, dv.labels = "Table 1: Linear Regression Model Results Estimating Crimes Per Week", pred.labels = c("Intercept", 'Median Income'))
```

@@

## Regression Between Median Income and Crimes per Zip Code

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center"}
lm_income_crimes <- lm(HomicideCount ~ cali_median_income, data = homicide_income_by_year)

@@

sjPlot::tab_model(lm_income_crimes, dv.labels = "Table 2: Linear Regression Model Results Estimating Crimes Based on Median Income", pred.labels = c("Intercept", 'Median Income'))
```

@@

```{r, warning=FALSE, message=FALSE, echo=FALSE, fig.align = "center"}
library(kableExtra)
crimes_by_cb <- read.csv("0_Data/Cleaned_Data/crimes_by_census_block.csv")
colnames(crimes_by_cb) <- c(colnames(crimes_by_cb)[-c(1, length(colnames(crimes_by_cb)))],"LON", "LAT")

@@

pretty_labels <- c("(Intercept)", paste("Census Block", 2:6))
sjPlot::tab_model(model, pred.labels = pretty_labels, dv.labels = "Table 3: Annual Homicide Count by Census Block", show.est = TRUE)
```

The results of this model show a significant relationship between Census Block and the homicide rate, except for Census Block 2. Census Block 2 experiences a higher incidence of homicides, but it is not significantly different. The other Census Blocks show a statistically significant decrease in homicide rates. However, the R2 might suggest overfitting.

## Linear Regression Models

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center"}
homicides_latlon_rounded <- homicides %>%
  mutate(LAT_rounded = round(LAT, 2),
         LON_rounded = round(LON, 2))

homicides_lat_long_yr <- homicides_latlon_rounded %>% 
  group_by(year, LAT_rounded, LON_rounded) %>%
  summarize(HomicideCount = n()) %>% ungroup()

linear_model1 <- lm(HomicideCount ~ LAT_rounded + LON_rounded, data = homicides_lat_long_yr)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center"}

linear_model2 <- lm(HomicideCount ~ year, data = homicides_lat_long_yr)

linear_model3 <- lm(HomicideCount ~ LAT_rounded + LON_rounded + year, data = homicides_lat_long_yr)

linear_model4 <- lm(HomicideCount ~ LAT_rounded * LON_rounded * year, data = homicides_lat_long_yr)

sjPlot::tab_model(linear_model1, linear_model2, linear_model3, linear_model4, pred.labels = c("(Intercept)", "Latitude", "Longitude", "Year", "Latitude:Longitude", "Latitude:Year", "Longitude:Year"), dv.labels = c("Model 1", "Model 2", "Model 3", "Model 4"), show.ci = FALSE, title = 
                    "Table 4: Geospatial Linear Regression Models Predicting Homicide Count")

Table 4 shows the results of four different linear regression models that were run using homicide counts by year and by latitude and longitude. Longitude and latitude were rounded to two decimal places to summarize the counts but maintain a good level of granularity in the coordinates data. Model 1 looked only at the geospatial aspect of the data and found that both latitude and longitude were significant predictors of homicide rate. As latitude increases, homicide rate decreases, meaning that homicides are more common towards the south and as you move north, homicide counts drop. As longitude increases, the number of homicides increases as well, meaning that homicides are more common to the East.

The second model finds that year is significant as a predictor of homicide rate, and a model looking at both spatial and temporal characteristics also finds year to be significant.

The fourth model, which also considered the interactions between these three features, did not find any significance.

@@

```{r DATA_Wrangling, output=FALSE, eval=FALSE}
library(sf) # for map
library(tidyverse) # using only for joining datasets to build map visual
library(RColorBrewer) # for map color scheme
library(tidygeocoder) # to get zip codes of addresses
library(dplyr)

@@

@@

```{r eval=FALSE, fig.cap="Crime densities in Los Angeles by LAPD Division, message=FALSE, warning=FALSE, 2010.2019 and 2020.Present", fig.align="center"}
lapd_shape <- read_sf(dsn = "0_Data/Raw_Data/LAPD_Divisions", layer = "LAPD_Divisions")

@@

```{r, eval=FALSE, warning=FALSE, message=FALSE, fig.cap = "Weekly Crime Counts in Los Angeles from 2010 to present", fig.align = "center"}
crime_la$DateParsed <- as.POSIXct(crime_la$Date.Rptd, 
                                  format = "%m/%d/%Y %I:%M:%S %p", 
                                  tz = "America/Los_Angeles")

@@

```{r, eval=FALSE, fig.cap = "Los Angeles Crimes by Day of the Week", warning=FALSE, message=FALSE, fig.align = "center"}
# Convert Date.Rptd to Date format (ignore time part)
#crime_la$Date.Rptd
crime_la$Date <- as.Date(substr(crime_la$Date.Rptd, 1, 10), format = "%m/%d/%Y")

@@

# Order the days correctly
day_order <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
crime_counts <- data.frame(crime_counts[day_order])
colnames(crime_counts) <- c("Day", "CrimeCount")

crime_counts$thousands <- crime_counts$CrimeCount/1000
